### 2015-03-19 manuscript/plos_template.tex
a number of modifications such as addition of method overview figures and many corrections of manuscript

Several statistical approaches for extracting characteristic mutation signatures have been proposed.
However, in previous approaches,
since the number of parameters increases exponentially as more contextual factors are taken into account,
we could not treat many contextual factors all together because of instability of estimation results.
Furthermore, interpretation of mutations signatures of huge dimensional vectors is often troublesome.

In this paper, we propose a novel approach based on hierarchical probabilistic modeling.
The proposed approach reduces the number of parameters by the independence assumption on each factor 
so that we can obtain more robust and interpretable estimates.
Using synthetic and real data, we demonstrate that the proposed approach can not only give highly robust estimates,
but also capture novel characteristics such as base frequency at the two base 5' to the mutated sites.
In addition, we clarify the relationships between the proposed approach and the ``mixed-membership models'', 
that have been actively studied in statistical machine learning and statistical genetics community.
Recognizing these relationships will help us to develop a grasp of mutation signature extraction problems,
and will allow us to utilize a lot of techniques accumulated on other fields to further improve the statistical methods.

Finally, we have prepared an R package of the proposed approach (probabilistic mutation signature, pmsignature),
which is available at \url{https://github.com/friend1ws/pmsignature}.

@@

It is known that the pattern of somatic mutations depends on cancer types and even individuals within the same caner type.
For example, C $>$ A mutations are frequent in lung caner
whereas C $>$ T and CC $>$ TT mutations are frequent in skin cancer,
and their patterns are usually associated to some kinds of carcinogens.
However, since the cost and throughput of sequencing technologies were limited, 
high-resolution analysis of mutation patterns was not possible.

With the coming of high-throughput sequencing technologies, we can now obtain a massive amount of somatic mutation data from cancer genomes,
and there is a great possibility of detecting novel mutation patterns leading to identification of novel carcinogens
and obtaining better classification of cancer genomes based on mutation patterns differences.
On the other hand, there is an increasing need for novel statistical method for mutation patterns analysis 
from vast amount of somatic mutation data.

@@

Cancer is a genomic disease. 
As we lead a life, DNAs within the cells throughout our body acquire a number of random somatic mutations
mainly caused by DNA replication errors and exposures to mutagens such as chemical substances, radioactivities and inflammatory reactions.  
Although most mutations are harmless (called ``passenger mutations''), 
a small portions of mutations at some specific sites in cancer genes (called ``driver mutations'') 
confer growth activities to the cells having them over other cells,
allowing such as autonomous proliferation and tissue invasion,
and contribute to oncogenesis \cite{stratton2009cancer}.
The main goal in cancer genome study has been to find the driver mutations to understand the mechanism of cancer development.
On the other hand,  even passenger mutations can also give us an important information, 
because they often show specific mutation patterns, or ``mutation signatures'', which reflect driving forces causing somatic mutations.
Classical analyses of mutation patterns have revealed a number of relationships between mutation patterns and carcinogens.
For example, C $>$ A mutations at CpG sites are abundant in lung cancers with smoking history,
and these are caused by benzo(a)pyrene included in tobacco smoke \cite{pmid12379884}.
Also, C $>$ T and CC $>$ TT  mutations are abundant in ultraviolet-light-associated skin cancers, 
and these are caused by pyrimidine dimers as a result of ultraviolet radiation \cite{pmid15748635}.

There were several problems in these classical studies \cite{pmid12379884, pmid15748635}.
Due to limited sequencing throughput, 
they mostly aggregated mutations collected from each individuals over the same cancer types focusing on a few cancer genes such as TP53
where high mutation frequencies could be expected,
and compared mutation patten profiles among different cancer types.
However, many of those mutations in those cancer genes are considered to be driver mutations adding selective activities to the cells,
which can give biased mutation profiles from those purely reflect driving forces of mutations.
Furthermore, the mutation pattern profile not for each cancer type but for each individual could not be explored in the above approach.

Nowadays, recent advancement in high-throughput sequencing technologies provides vast amounts of passenger mutations
and great chances for not only identification of novel driver mutations but also investigation of sample-by-sample mutation patterns in an unbiased way. 
A study using 21 breast cancer samples identified the association between C $>$ [AGT] mutations at TpC sites, 
which is later proved to be caused by APOBEC proteins, and the novel phenomenon called {\it kataegis} \cite{pmid22608084}. 
Moreover, the recent landmark study revealed a landscape of mutation signatures using from 7,034 primary cancer samples of 30 different classes \cite{pmid23945592}. 
Furthermore, it is expected that detection of novel mutation signatures and associated mutagens can lead to identification of novel mutagens and prevention of cancer.

At the same time, for extracting prominent mutation signatures from vast amounts of somatic mutation data,
there is an increasing need for novel latent variable modeling approaches.
Currently, only a few approaches have been proposed. 
In \cite{pmid23318258}, nonnegative matrix factorization (NMF, \cite{pmid10548103}) is used for the matrix whose elements represent frequencies of mutation patterns for each sample. 
In \cite{pmid23628380}, the number of each mutation patterns is assumed to be generated by Poisson distribution whose parameters are linear combinations of latent mutational signatures.

One of the major problems of the previous approach is that 
the number of parameters in the model grows exponentially 
with the number of contextual factors to take into account.
Many of recent approaches take into account just immediate 5' and 3' bases as contextual factors in addition to substitution patterns.
However, when considering two 5' and 3' bases to the mutated sites,
the estimated mutation signatures often become unstable due to the curse of high dimensionality, 
despite there is a potential need to observe two base 5' and 3' positions to the mutated sites \cite{pmid9683596}.
Also, it is often very hard to interpret and extract important characteristics from the estimate with high-dimensional parameters.
Furthermore, considering that latent variable modeling have enormous number of researches in statistics and machine learning, 
there is a great possibility that we can further elaborate statistical methods for mutational signature extraction problems.

In this paper, we present a novel probabilistic approach.
To avoid the problem of the exponential increase of the number of parameters,
we assumed independence among contextual factors, which can give more interpretable and robust estimation.
We demonstrate that independence assumption can robustly capture known mutational signatures with additional contextual informations.

In addition, even though it has not been widely discussed, 
the problem of identifying mutation signatures is closely related to "mixed-membership models'' in other fields, 
such as ``population admixture'' models \cite{pmid10835412} in the analysis of population structure, 
and the ``latent dirichlet allocation'' (LDA) model for document clustering \cite{Blei:2003}.
In this paper, we explicitly show the relationships between the proposed model and the mixed-membership models and discuss the relationships,
which we believe will be highly helpful for future elaboration of the statistical method for mutation signature extraction problems.

@@

One of the major problems is that 
total number of mutation patterns exponentially increases
as we increase the number of mutation features to take into account.
For example, if we want to take account of up to $n$ bases 5' and 3' to the mutated site 
(which we call $-n$ site and $+n$ site, respectively, in this paper) as well as 6 substitution patterns, 
the number of parameters becomes $6 \times 4^{2n} - 1$. 
On the other hand, the mutation rates are mostly below 10 per Mb \cite{pmid23945592},
(where about 30,000 mutations are expected in entire genomic regions).
Therefore, considering even $\pm 2$ sites and substitution patterns (1535 parameters)
is often difficult because of instability of parameter estimation.

On the other hand, observing the mutation signatures observed in the previous studies, 
many mutation signatures had clear characteristics such as:
\begin{itemize}
\item 
A certain value of mutation features is often dominant in many known mutation signatures. 
For example, the base T is highly dominant at the $-1$ site in the APOBEC signature, 
and C $>$ T substitution is highly dominant in the ultraviolet signature \cite{pmid23945592} (see Figure \ref{APOBEC_nature2013_sig}).

\item
Proportional relationships among mutation features are often observed in many signatures as shown in the APOBEC signature, 
where the frequencies of bases at the $+1$ site are largely proportional across 3 major substitution patterns C$>$T, C$>$G and C$>$A,
and show consistent tendencies (A and T bases are more frequent than C and G bases).

@@

In this paper, we propose to represent mutation signatures with independent multiple multi-nominal distributions
on each mutation feature. 
By doing this, we can reduce the number of parameters from exponential to linear with respect to the number of mutation features. 
When we consider up to $\pm n$ sites as well as 6 substitution patterns, 
the required number of parameters becomes $5 + 6n$. 
For example, when considering $\pm 2$ sites and substitution patterns,
the number of parameters becomes 17, which is far less compared to 1535 in the previous approach.
Furthermore, independence assumption enables us to come up with fairly interpretable representation for mutation signatures.

@@

\begin{enumerate}
\item
For representing the transcription factor binding motifs, the position specific weight matrix (PSWM), 
which is equivalent to independent multiple multinomial distributions, 
has been quite successfully utilized 
even though it is never likely that transcription factors independently choose the bases of their binding sites. 
This is probably because PSWM can capture important characteristics of transcription factor binding sites 
in spite of the independence assumption,
and can be represented in an intuitively interpretable way via sequencing logos \cite{pmid2172928}.

\item
As we demonstrate in later sections, 
the proposed approach can robustly extract most of previously-collected mutation signatures even independence assumption is imposed.

\item
Although most mutation signatures show proportional relationships among mutation features, 
a few signatures seem to violate proportional relationships and cannot be explained by independent models. 
For example, the pol $\epsilon$ mutation signature \cite{pmid23945592} (see Figure \ref{POLE_nature2013_sig}) puts strong probability masses 
on just two pattern TpCpT $>$ TpApT and TpCpG $>$ TpTpG, 
and the frequencies of substitution patterns and $+1$ bases are not proportional unlike the APOBEC signature.
However,  by using multiple mutation signatures, we can represent this phenomenon even with independence assumption.
\end{enumerate}

@@

The proposed method has mainly two novelties compared to existing approaches.
The first point is the assumption of independence, where we can identify mutation signatures as multiple multinomial distribution parameters on multiple contextual factors.
We give a way of visualizing probabilistic mutation signature (see Figure \ref{mutSig_example}), 
which is reminiscent of sequencing logos \cite{pmid2172928}. 
This representation will help us to immediately identify which elements are strongly working on each contextual factors.
In this paper, we focus on substitution patterns, flanking bases (up to the $\pm 2$ positions) and transcribed directions as contextual factors of mutation patterns.

The second novel point is that we adopt  the ``mixed membership model'' for the mutation signature problem.
Briefly, the mixed membership model is basically clustering method, but allowing that each individual can have multiple clusters.
The mixed membership model is widely used in many different fields such as population structure classification in population genetics and document classification problems in machine learning.
This model is also shown to be closely related to nonnegative matrix factorization.

Suppose there are in total $K$ source of mutations (or mutation signatures) in the cohort of interest.
For each cancer genome, somatic mutations are generated by one of the $K$ sources,
and the ratio of the amount of contributing sources varies across cancer genomes depending on the ages, lifestyle, races and so on.
We call this parameter as ``membership parameters.''

Here, by convention of the mixed membership model, we assume that each mutation for each cancer genome is generated by a two step model:
First, one of the contributing mutation signature is chosen depending on the membership parameter of that cancer genome.
Then, according to the probability distribution of the selected mutation signature, 
mutation features such as substitution patterns and flanking base pairs are generated.
Although the mutation signatures and membership parameters are unknown,
we can efficiently estimate these parameters by an EM-algorithm.

To account for the intrinsic sequential context, we add background signatures.
See the Method section for the mathematical details, 
detailed relationships with mixed membership models and nonnegative matrix factorizations, how to choose the number of mutation signatures and so on.

@@

First, we would like to investigate whether the proposed approach can extract ``true'' mutation signatures or not.
Since we can not know the true mutation signatures in real biological data, we resorted to simulation studies.

Here, we generated a set of mutations changing the number of cancer genomes (10, 25, 50, 100), 
and the number of mutations for each cancer genome (10, 25, 50, 100, 250, 500, 1000).
The number of mutation signatures were set to 5 including background mutation ratio.
The mutation feature parameters and signature distribution parameters were generated by Dirichlet distribution,

\begin{equation*}
 f_{k,l,m_l} \sim \text{Dir} (\alpha \bm{1} ),\ k = 1, \cdots, K,\  l = 1, \cdots, L.
\end{equation*}
 \begin{equation*}
 q_{i, k} \sim \text{Dir} (\gamma \bm{1} ),\ i = 1, \cdots, I,
 \end{equation*}
where the $\alpha$ and $\gamma$ represent the amounts of dispersion 
for the mutation features parameters and signature distribution parameters, respectively.
When these values are smaller, only a fewer components can have larger probability masses.
and the rest will have much smaller masses.  
When these are large, all the components tend to have evenly-distributed probability masses.

As Figure \ref{sim_facet} shows, we could estimate the mutation signatures very accurately overall.
(see Supplementary Figure 1 for an example).
In most cases, the log-likelihood stopped increasing at five mutation signatures, 
whereas the standard error of the estimated parameters started increasing past five mutation signatures (see Supplementary Figure 2),
indicating that true number of mutation signatures can be recovered by observing the trade-off between likelihood and standard-errors at least in an ideal case.

As expected, as we increase the number of either cancer genomes or mutations, the accuracy of the estimates increased.
Also, as we increase the value of $\alpha$, then the accuracy of the estimated signatures got worse.
However, even we increase the value of $\gamma$, the accuracy of the estimate signatures did not changed much.
Therefore, the dispersion of mutation feature parameters influences the accuracy of mutation signatures
more sensitively than signature distribution parameters.

@@

In this section, we compare the ``full model'' and the ``independent model'' by examining mutation signatures obtained by the two approaches,
and investigate the robustness by downsampling experiments.
The dataset used here is a list of 14717 somatic substitutions 
collected from the study of 26 urothelial carcinomas of the upper urinary tract (UTUC) \cite{pmid23926200},
where they found a novel mutation signature: 
T $>$ A substitutions at CpTpG sites with a strong strand specificity caused by aristolochic acids (AA).

We performed the proposed method considering substitution patterns, up to $\pm 2$ sites and strand directions as mutation features.
For the full model, we assign one integer for each combination of these features, and thus $L = 1, \bm{M} = (3072)$.
For the independent model, $L = 6, \bm{M} = (6, 4, 4, 4, 4, 2)$.

First, performing on various number of mutation signatures $K$ (including a background signature) on the independent model, 
we observed how the detected mutation signatures change (See Supplementary Figure 3).
For $K = 2$, a mutation signature which seem to correspond to AA  (T $>$ A substitutions at CpTpG sites with strong transcription direction) was observed (Supplementary Figure 3 (a)).
For $K = 3$, an additional mutation signature corresponding APOBEC enzyme (C $>$ [AGT] at TpCpN sites) was observed
in addition to the AA mutation signature (Figure \ref{UTUC:APOBEC_ind5_sig}, \ref{UTUC:AA_ind5_sig}, Supplementary Figure 3 (b)).
For $K = 4$, an additional signature (T $>$ A at NpTpN sites with strong transcription direction) 
that is somewhat similar to the AA signature was observed (Supplementary Figure 3 (c)). 
When we checked the correlation of estimated membership parameter for each cancer genome, 
strong correlation between the AA signature (CpTpG $>$ CpApG) and that additional AA-like signature could be observed ($R = 0.77$, see Supplementary Figure 5). 
This additional signature may be just making up for the residual of the AA signature 
which the original AA signature could not explain due to a slight deviance of the probabilistic model.
The values of likelihood continued increasing as the $K$ increase, while the bootstrap-errors started to increase at $K = 5$ (See Supplementary Figure 4).
The strong correlation among estimated membership parameters started to be shown at $K = 4$.
Considering all these factors together, $K = 3$ seems to be a reasonable choice in terms of the interpretability,
and we adopted $K = 3$ in the following.

For the independent model, we could observed the depletion of G base at the $-2$ site,
which is consistent to the previous study \cite{pmid23318258} and the result in the next subsection.
On the other hand, for the full representation model, this tendency was rather mild (Figure \ref{UTUC:APOBEC_full5_sig}, \ref{UTUC:AA_full5_sig}, Supplementary Figure 6).
Inferred AA mutation signature had no clear characteristics at two bases 5' and 3' to the mutated site compared to the APOBEC mutation signature.

Assuming that the mutation signatures obtained using whole 14717 substitutions as a gold standard, 
we performed the proposed method on down-sampled data (1\%, 2.5\%, 5\%, 10\%, 25\%, 50\%), 
and compared obtained signatures with the gold standard.
For measuring the deviations of the mutation signature, the cosine similarity was used on the $\prod_{l=1}^L M_l$ dimensional vector space $f_{i, \bm{m}} = \prod_{l=1}^L f_{k,l,m_l}$,
so that the comparison between the full model and the independent model become possible.
Trials for downsampling experiment for each ratio and model were repeated for 100 times.

As the Figure \ref{UTUC:APOBEC_downsampling} and \ref{UTUC:AA_downsampling} shows, 
the independent representation could recover the original signatures even when, e.g, about 90 \% of the data was removed.
In this experiment, the AA signature was more robust than the APOBEC signature.
We believe that this is because the number of T $>$ A substitutions at GpTpC sites are far more frequent in this dataset.
These results indicate that the independent model can give more robust estimates than the full model.

@@

The signature 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
are arguably associated with deregulated activity of the error-prone polymerase Pol $\epsilon$.
The proposed method uses two signatures for representing the signature for the Pol $\epsilon$ dysfunction, 
while was represented by one signature in the previous study.
Interestingly, the signature 1 shows transcription biases whereas the signature 8 does not.
This phenomena are consistent for both colorectal and uterus cancers (Figures \ref{POLE1_sig_list} and \ref{POLE2_sig_list}).
C $>$A at TpCpT mutations were shown to be enriched in leading strands of replication forks.
whereas, there is no mention about the replication strand-specificity of C $>$ T at TpCpG mutations \cite{pmid25228659}.
Although the replication strand bias is different from the transcription strand bias, these two biases may have strong correlation
through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.
Therefore, the transcription strand bias observed in the signature 1 may actually come from the replication strand biases.

>>>>

Several statistical approaches for extracting characteristic mutation signatures have been proposed.
However, in previous approaches,
since the number of parameters increases exponentially as more contextual factors are taken into account,
we could not treat many contextual factors all together because of instability of estimation results.
Furthermore, interpretation of mutations signatures with huge dimensional vectors is often troublesome.

In this paper, we propose a novel approach based on hierarchical probabilistic modeling.
The proposed approach reduces the number of parameters by an independence assumption on each factor 
so that we can obtain more robust and interpretable estimates.
Using synthetic and real data, we demonstrate that the proposed approach can not only give highly robust estimates,
but also capture novel characteristics such as base frequency at the two base 5' to the mutated sites.
In addition, we clarify the relationships between the proposed approach and the ``mixed-membership models'', 
that have been actively studied in statistical machine learning and statistical genetics community.
Recognizing these relationships will help us to develop a grasp of mutation signature extraction problems,
and will allow us to utilize a number of techniques accumulated on other fields to further improve the statistical methods.

We have prepared an R package of the proposed approach (probabilistic mutation signature, pmsignature),
which is available at \url{https://github.com/friend1ws/pmsignature}.

@@

The pattern of somatic mutations has been known to depend on cancer types and even individuals within the same caner type.
For example, C $>$ A mutations are frequent in lung caners
whereas C $>$ T and CC $>$ TT mutations are frequent in skin cancers,
and their patterns are usually associated to some kinds of carcinogens such as tobacco smoking and ultraviolets.
However, since the cost and throughput of sequencing technologies were limited, 
high-resolution analyses of mutation patterns were not possible.

With the coming of high-throughput sequencing technologies, we can now obtain a massive amount of somatic mutation data from cancer genomes,
and there is a great possibility of detecting novel mutation patterns leading to identification of novel carcinogens
and better classification of cancer genomes.
On the other hand, there is an increasing need for novel statistical method for analyzing mutation patterns
from vast amounts of somatic mutation data.

@@

Cancer is a genomic disease. 
As we lead a life, DNAs within the cells throughout our body acquire a number of random somatic mutations
mainly caused by DNA replication errors and exposures to mutagens such as chemical substances, radioactivities and inflammatory reactions.  
Although most mutations are harmless (called ``passenger mutations''), 
a small portions of mutations at some specific sites in cancer genes (called ``driver mutations'') 
confer growth activities to the cells over other cells,
causing autonomous proliferation, tissue invasion, and contribute to oncogenesis \cite{stratton2009cancer}.
The main goal in cancer genome study has been to find driver mutations to understand the mechanism of cancer development.
On the other hand,  even passenger mutations can also give us an important information, 
because they often show specific mutation patterns, or ``mutation signatures'', which reflect driving forces causing somatic mutations.
Classical analyses of mutation patterns have revealed a number of relationships between mutation patterns and carcinogens.
For example, C $>$ A mutations at CpG sites are abundant in lung cancers with smoking history,
and these are caused by benzo(a)pyrene included in tobacco smokes \cite{pmid12379884}.
Also, C $>$ T and CC $>$ TT  mutations are abundant in ultraviolet-light-associated skin cancers, 
and these are caused by pyrimidine dimers as a result of ultraviolet radiation \cite{pmid15748635}.

There were several problems in these classical studies \cite{pmid12379884, pmid15748635}.
Due to limited sequencing throughput, 
they mostly aggregated mutations collected from each individual over the same cancer type focusing on a few cancer genes such as TP53
where high mutation frequencies could be expected,
and compared mutation patten profiles among different cancer types.
However, since many of the mutations in cancer genes are considered to be driver mutations adding selective activities to the cells,
the resultant mutation profiles are considered to biased from those purely reflect driving forces of mutations.
Furthermore, the mutation patterns not for each cancer type but for each individual could not be explored in the above approach.

Nowadays, recent advancement in high-throughput sequencing technologies provides vast amounts of passenger mutations
and great chances for not only identification of novel driver mutations but also investigation of sample-by-sample mutation patterns in an unbiased way. 
A study using 21 breast cancer samples identified the association between C $>$ [AGT] mutations at TpC sites, 
which is later proved to be caused by APOBEC protein family, and a novel phenomenon called {\it kataegis} \cite{pmid22608084}. 
Moreover, a landmark study using from 7,034 primary cancer samples of 30 different classes revealed an overview of mutation signatures in most cancer types \cite{pmid23945592}. 
Now, there are great hopes that detection of novel mutation signatures and associated mutagens can lead to identification of novel mutagens and prevention of cancer.

At the same time, for extracting prominent mutation signatures from vast amounts of somatic mutation data,
there is an increasing need for novel statistical approaches.
Currently, only a few approaches have been proposed. 
In \cite{pmid23318258}, nonnegative matrix factorization (NMF, \cite{pmid10548103}) is used for the matrix whose elements represent frequencies of mutation patterns for each sample. 
In \cite{pmid23628380}, the number of each mutation patterns is assumed to be generated by Poisson distribution whose parameters are linear combinations of latent mutational signatures.

One of the major problems of the previous approaches is that 
the number of parameters in the model grows exponentially 
with the number of contextual factors to take into account.
Many of recent approaches take into account just immediate 5' and 3' bases as contextual factors in addition to substitution patterns.
However, when considering two 5' and 3' bases to the mutated sites,
the estimated mutation signatures often become unstable due to the curse of high dimensionality, 
despite there is a potential need to observe two base 5' and 3' positions to the mutated sites \cite{pmid9683596}.
Also, it is often very hard to interpret the estimates with high-dimensional parameters.

In this paper, we present a novel probabilistic approach based on ``mixed membership models,'' that is widely used in other fields such as population genetics and machine learning.
Even though it has not been widely discussed, there is close relationships between the mutation signature problems and mixed membership models,
and we believe that being aware of those relationships will be very helpful for future elaboration of the statistical methods.
Furthermore, to avoid the problem of the exponential increase of the number of parameters,
we assume an independence among contextual factors, which can give more interpretable and robust estimation.
We demonstrate that the independence assumption can robustly capture known mutational signatures with additional contextual informations.

% In addition, even though it has not been widely discussed, 
% the problem of identifying mutation signatures is closely related to "mixed-membership models,''
% that is widely used in other field such as ``population admixture'' models \cite{pmid10835412} in the analysis of population structure, 
% and the ``latent dirichlet allocation'' (LDA) model for document clustering \cite{Blei:2003}.
% In this paper, we explicitly show the relationships between the proposed model and the mixed-membership models and discuss the relationships,
% which we believe will be highly helpful for future elaboration of the statistical method for mutation signature extraction problems.

@@

One of the major problems is that 
total number of mutation patterns exponentially increases
as we increase the number of mutation features to take into account.
For example, if we want to take account of up to $n$ bases 5' and 3' to the mutated site 
(which we call $-n$ position and $+n$ position, respectively, in this paper) as well as 6 substitution patterns, 
the number of parameters becomes $6 \times 4^{2n} - 1$. 
On the other hand, the mutation rates are mostly below 10 per Mb \cite{pmid23945592},
(where about 30,000 mutations are expected in entire genomic regions).
Therefore, considering even $\pm 2$ positions and substitution patterns (1535 parameters)
is often difficult because of instability of parameter estimation.

On the other hand, observing the mutation signatures detected in the previous studies, 
many mutation signatures have clear characteristics such as:
\begin{itemize}
\item 
A certain value of mutation features is often dominant in many known mutation signatures. 
For example, the base T is highly dominant at the $-1$ position in the APOBEC signature, 
and C $>$ T substitution is highly dominant in the ultraviolet signature \cite{pmid23945592} (see Figure \ref{APOBEC_nature2013_sig}).

\item
Proportional relationships among mutation features are often observed in many signatures as shown in the APOBEC signature, 
where the frequencies of bases at the $+1$ position are largely proportional across 3 major substitution patterns C$>$T, C$>$G and C$>$A,
showing consistent tendencies  (A and T bases are more frequent than C and G bases).

@@

In this paper, we propose to represent mutation signatures with independent multiple multi-nominal distributions
on each mutation feature,
and thus, we can reduce the number of parameters from exponential to linear with respect to the number of mutation features. 
When we consider up to $\pm n$ positions as well as 6 substitution patterns, 
the required number of parameters becomes $5 + 6n$. 
For example, when considering $\pm 2$ positions and substitution patterns,
the number of parameters becomes 17, which is far less compared to 1535 in the previous approach without the independence assumption.
Furthermore, independence assumption enables us to come up with fairly interpretable representation for mutation signatures.

@@

\begin{enumerate}
\item
For representing the transcription factor binding motifs, the position specific weight matrix (PSWM), 
which is equivalent to independent multiple multinomial distributions, 
has been quite successfully utilized 
even though it is never likely that transcription factors independently choose the nucleotides of their binding sites. 
The success of the PSWM may be because it can capture important characteristics of transcription factor binding sites 
in spite of the independence assumption,
and can be represented in an intuitively interpretable way via sequencing logos \cite{pmid2172928}.

\item
As we demonstrate in later sections, 
the proposed approach can robustly extract most of previously-collected mutation signatures even independence assumption is imposed,
indicating the practical validity of the independence assumption.

\item
Although most mutation signatures show proportional relationships among mutation features, 
a few signatures seem to violate proportional relationships and cannot be explained by independent models. 
For example, the pol $\epsilon$ mutation signature \cite{pmid23945592} (see Figure \ref{POLE_nature2013_sig}) puts strong probability masses 
on just two pattern TpCpT $>$ TpApT and TpCpG $>$ TpTpG, 
and the frequencies of substitution patterns and bases at the $+1$ positions are not proportional unlike the APOBEC signature.
However, by using multiple mutation signatures, we can represent this phenomenon even under the independence assumption.
\end{enumerate}

@@

The proposed method has mainly two novelties compared to existing approaches.
The first point is, as noted above, the assumption of independence, where we can identify mutation signatures as multiple multinomial distribution parameters on multiple contextual factors.
We give a way of visualizing probabilistic mutation signature (see Figure \ref{mutSig_example}), 
which is somewhat reminiscent of sequencing logos \cite{pmid2172928}. 
This representation will help us to immediately identify which elements are strongly expressed on each mutation factors.
In this paper, we focus on substitution patterns, flanking bases (up to the $\pm 2$ positions) and transcribed directions as contextual factors of mutation patterns.

The second point is that we adopt the framework of ``mixed membership models'' for the mutation signature problem.
The mixed memberships model are basically clustering methods, but allowing that each individual can have multiple clusters,
and widely used in many different fields such as population structure classification in population genetics \cite{pmid10835412} and document classification problems in machine learning \cite{Blei:2003}.
This models are also shown to be closely related to nonnegative matrix factorization \cite{ding2008equivalence}.

The generative process of somatic mutations proposed proposed in this paper is summarized in the Figure \ref{methodOverview}.
Suppose there are in total $K$ sources of somatic mutations (or mutation signatures) in the cohort of interest.
For each cancer genome, somatic mutations are generated by one of the $K$ sources,
and the ratio of the amount of contributing sources varies across cancer genomes depending on the ages, lifestyle, races and so on.
We call this parameter as ``membership parameters.''
Here, by convention of the mixed membership models, we assume that each mutation for each cancer genome is generated by a two step model.
First, one of the contributing mutation signatures is chosen depending on the membership parameter of the corresponding cancer genome.
Then, according to the probability distribution of the selected mutation signature, 
mutation features such as substitution patterns and flanking bases are generated.
Although the mutation signatures and membership parameters are unknown,
we can efficiently estimate these parameters by an EM-algorithm.

To account for the intrinsic sequential context, we add background signatures calculated by compositions of short nucleotides sequences in the human genome.
See the Method section for the mathematical details, 
detailed relationships with mixed membership models and nonnegative matrix factorizations, parameter estimation details, how to choose the number of mutation signatures and so on.

@@

First, we would like to investigate whether the proposed approach can extract ``true'' mutation signatures or not.
Since we cannot know the true mutation signatures in real biological data, we resorted to simulation studies.

Here, we assumed substitution pattens and up to $\pm 2$ flanking bases as mutation features,
and generated a set of mutations changing the number of cancer genomes (10, 25, 50, 100), 
and the number of mutations for each cancer genome (10, 25, 50, 100, 250, 500, 1000).
The number of mutation signatures were set to 5 including background mutation ratio.
The mutation feature parameters and membership parameters were generated by Dirichlet distribution,

\begin{equation*}
 f_{k,l,m_l} \sim \text{Dir} (\alpha \bm{1} ),\ k = 1, \cdots, K,\  l = 1, \cdots, L.
\end{equation*}
 \begin{equation*}
 q_{i, k} \sim \text{Dir} (\gamma \bm{1} ),\ i = 1, \cdots, I,
 \end{equation*}
where the $\alpha$ and $\gamma$ represent the amounts of dispersion 
for the mutation signature parameters and membership parameters, respectively.
When these values are smaller, only a fewer components can have larger probability masses and the rest will have much smaller masses.
On the other hand, when these are large, all the components tend to have evenly-distributed probability masses.

As Figure \ref{sim_facet} shows, we could estimate the mutation signatures very accurately overall (see Supplementary Figure 1 for an example).
In most cases, the log-likelihood stopped increasing at five mutation signatures, 
whereas the standard error of the estimated parameters started increasing past five mutation signatures (see Supplementary Figure 2),
indicating that true number of mutation signatures can be recovered by observing the trade-off between likelihood and standard-errors at least in an ideal case.

As expected, as we increase the number of either cancer genomes or mutations, the accuracy of the estimates increased.
Also, as we increase the value of $\alpha$, then the accuracy of the estimated signatures became worse.
However, even we increase the value of $\gamma$, the accuracy of the estimate signatures did not changed much.
Therefore, the dispersion of mutation feature parameters influences the accuracy of mutation signatures
more sensitively than that of membership parameters.

@@

In this subsection, we compare  the ``independent model,'' where the independence assumption are imposed among mutation features
and the ``full model,'' where no independence assumption are imposed (see Method section for the mathematical detail)
by examining mutation signatures obtained by the two approaches and investigating the robustness by downsampling experiments.
The dataset used here is a list of 14717 somatic substitutions 
collected from the study of 26 urothelial carcinomas of the upper urinary tract (UTUC) \cite{pmid23926200},
where they found a novel mutation signature: 
T $>$ A substitutions at CpTpG sites with a strong strand specificity caused by aristolochic acids (AA).

We performed the proposed method considering substitution patterns, up to $\pm 2$ positions and strand directions as mutation features.
For the full model, the number of parameters per mutation signature is 3071, where as it is 18 for the independent model.

First, performing on various number of mutation signatures $K$ (including a background signature) on the independent model, 
we observed how the detected mutation signatures change (see Supplementary Figure 3).
For $K = 2$, a mutation signature which seem to correspond to AA  (T $>$ A substitutions at CpTpG sites with strong transcription strand specificity) was observed (Supplementary Figure 3 (a)).
For $K = 3$, an additional mutation signature corresponding APOBEC enzyme (C $>$ [AGT] at TpCpN sites) was observed
in addition to the AA mutation signature (Figure \ref{UTUC:APOBEC_ind5_sig}, \ref{UTUC:AA_ind5_sig}, Supplementary Figure 3 (b)).
For $K = 4$, an additional signature (T $>$ A at NpTpN sites with strong strand specificity) 
that is somewhat similar to the AA signature was observed (Supplementary Figure 3 (c)). 
When we checked the correlation of estimated membership parameter for each cancer genome, 
strong correlation between the AA signature (CpTpG $>$ CpApG) and that additional AA-like signature could be observed ($R = 0.77$, see Supplementary Figure 5). 
This additional signature may be just making up for the residual of the AA signature 
which the original AA signature could not explain due to a slight deviance of the probabilistic model.
The values of likelihood continued increasing as the $K$ increase, while the bootstrap-errors started to increase at $K = 5$ (See Supplementary Figure 4).
The strong correlation among estimated membership parameters started to be shown at $K = 4$.
Considering all these factors together, $K = 3$ seems to be a reasonable choice in terms of the interpretability,
and we adopted $K = 3$ in the following.

For the independent model, we could observed the depletion of G base at the $-2$ position,
which is consistent to the previous study \cite{pmid23318258} and the result in the next subsection.
On the other hand, for the full model, this tendency was rather mild (Figure \ref{UTUC:APOBEC_full5_sig}, \ref{UTUC:AA_full5_sig}, Supplementary Figure 6).
Inferred AA mutation signature had no clear characteristics at the $-2$ position compared to the APOBEC mutation signature.

Assuming that the mutation signatures obtained using whole 14717 substitutions as a gold standard, 
we performed the proposed method on down-sampled data (1\%, 2.5\%, 5\%, 10\%, 25\%, 50\%), 
and compared obtained signatures with the gold standard.
For measuring the deviations of the mutation signature, the cosine similarity was used on the dimensional vector space full model,
so that the comparison between the full model and the independent model become possible.
Trials for downsampling experiment for each ratio and model were repeated for 100 times.

% For measuring the deviations of the mutation signature, the cosine similarity was used on the $\prod_{l=1}^L M_l$ dimensional vector space $f_{i, \bm{m}} = \prod_{l=1}^L f_{k,l,m_l}$,

As the Figure \ref{UTUC:APOBEC_downsampling} and \ref{UTUC:AA_downsampling} shows, 
the independent model could mostly recover the original signatures even when, e.g, about 90 \% of the data was removed,
whereas the full model could not recover, especially, the APOBEC signature.
In this experiment, the AA signature was more robust than the APOBEC signature.
We believe that this is because the number of T $>$ A substitutions at GpTpC sites are far more frequent in this dataset.
These results indicate that the independent model can give more robust estimates than the full model.

@@

The signature 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
are arguably associated with deregulated activity of the error-prone polymerase Pol $\epsilon$.
In the proposed method, the signature for the Pol $\epsilon$ dysfunction was represented by two signatures
while it was represented by one signature in the previous study.
Interestingly, the signature 1 shows transcription biases whereas the signature 8 does not.
This phenomena are consistent for both colorectal and uterus cancers (Supplementary Figures 7(c, d)).
C $>$A at TpCpT mutations were shown to be enriched in leading strands of replication forks,
whereas there is no mention about the replication strand-specificity of C $>$ T at TpCpG mutations \cite{pmid25228659}.
Although the replication strand bias is a different phenomenon from the transcription strand bias, these two biases may have strong correlation
through the fact that replication origins prefer transcription start sites \cite{pmid23187890},
and the transcription strand bias observed in the signature 1 may actually come from the replication strand biases.
